---
title: "Data Cleaning"
author: "group"
date: "September 17, 2019"
output:
  pdf_document: default
  html_document:
    df_print: paged
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(tidy.opts=list(width.cutoff=60),tidy=TRUE)
```

# Project on Scraping and Cleaning Data

This data is acquired from experiments have been carried out with a group of 30 volunteers within an age bracket of 19-48 years. Each person performed six activities (WALKING, WALKING_UPSTAIRS, WALKING_DOWNSTAIRS, SITTING, STANDING, LAYING) wearing a smartphone (Samsung Galaxy S II) on the waist. Using its embedded accelerometer and gyroscope, we captured 3-axial linear acceleration and 3-axial angular velocity at a constant rate of 50Hz. The experiments have been video-recorded to label the data manually. The obtained dataset has been randomly partitioned into two sets, where 70% of the volunteers was selected for generating the training data and 30% the test data. 

The sensor signals (accelerometer and gyroscope) were pre-processed by applying noise filters and then sampled in fixed-width sliding windows of 2.56 sec and 50% overlap (128 readings/window). The sensor acceleration signal, which has gravitational and body motion components, was separated using a Butterworth low-pass filter into body acceleration and gravity. The gravitational force is assumed to have only low frequency components, therefore a filter with 0.3 Hz cutoff frequency was used. From each window, a vector of features was obtained by calculating variables from the time and frequency domain. See 'features_info.txt' for more details. 

Information on the data may be found:

https://archive.ics.uci.edu/ml/datasets/human+activity+recognition+using+smartphones#

## Preparing workspace

This cleans up the workspace, removing exiting items:

```{r }
rm(list = ls())
```

## Getting and loading the required packages

This loads the required packages for the data cleaning process including a condition to install the packages if they are not already installed:

```{r, warning=FALSE}

if ("dplyr" %in% row.names(installed.packages()) == FALSE){install.packages("dplyr")}
if ("tidyr" %in% row.names(installed.packages()) == FALSE){install.packages("tidyr")}

library("dplyr"); 
library("tidyr")

```


## Extracting data from the web

Here the data is sourced from the website and stored. A check is also done to know if the data already exists.
```{r}

file_url <- "https://d396qusza40orc.cloudfront.net/getdata%2Fprojectfiles%2FUCI%20HAR%20Dataset.zip"

if (!dir.exists("data2")){
  dir.create("data2")
}

data_path = "./data2/UCI HAR Dataset"

if (!file.exists(data_path)){
  download.file(file_url, "./data2/UCI_data.zip")
  unzip("./data2/UCI_data.zip", exdir = "./data2")
  
}

```


## Merging the test and train datasets to create one dataset

The dataset includes the following files:


- 'README.txt'

- 'features_info.txt': Shows information about the variables used on the feature vector.

- 'features.txt': List of all features.

- 'activity_labels.txt': Links the class labels with their activity name.

- 'train/X_train.txt': Training set.

- 'train/y_train.txt': Training labels.

- 'test/X_test.txt': Test set.

- 'test/y_test.txt': Test labels.

The different files are read in and training and test sets are merged for subsequent cleaning:

```{r}

list.files("./data2/UCI HAR Dataset/")

#reading in train set values
x_train <- read.table(file = "./data2/UCI HAR Dataset/train/X_train.txt")
y_train <- read.table(file = "./data2/UCI HAR Dataset/train/y_train.txt", col.names = c("activity"))
sub_train <- read.table(file = "./data2/UCI HAR Dataset/train/subject_train.txt", col.names = c("subject"))

#combining the columns to make training data
dt_train <- cbind(sub_train, y_train, x_train)

#reading in the test set values
x_test <- read.table(file = "./data2/UCI HAR Dataset/test/X_test.txt")
y_test <- read.table(file= "./data2/UCI HAR Dataset/test/y_test.txt", col.names = c("activity"))
sub_test <- read.table(file= "./data2/UCI HAR Dataset/test/subject_test.txt", col.names = c("subject"))

#combinig the columns to make the training data
dt_test <- cbind(sub_test, y_test, x_test)

#merging test and train data sets
dt_merged <- rbind(dt_train, dt_test)

# displaying a sample data entry
print(head(dt_merged, 1))

```


## Extracting the feature names

The feature names are extracted from the relevant file here and additional feature names (`"subject", "activity"`) are included:

```{r}
features <- read.table(file = "./data2/UCI HAR Dataset/features.txt", stringsAsFactors = FALSE, col.names = c("id", "name"))
feature_extended <- c("subject", "activity", features$name)

# displaying some of the features
print(feature_extended[1:20])
```



## Extracting only measurements on mean and standard deviation

A decision is made here to extract only features (columns) which includes `mean,std,subject,activity` in the naming while dropping those which include `freq`. This is an arbitrary choice to simplify the data:

```{r}
feature_select <- feature_extended[grepl("mean|std|subject|activity",   feature_extended) & !grepl("Freq|freq", feature_extended)]
feature_select_index <- grepl("mean|std|subject|activity", feature_extended) &   !grepl("Freq|freq", feature_extended)

#filtering the needed variables in merged dataset by the position of the required features
dt_merged_filtered <- dt_merged[, feature_select_index]

head(dt_merged_filtered, 1)

```


## Cleaning up variable names

The activity labels are extracted from the relevant file and merged with the data:

```{r}
#reading in activity names
activity_labels <- read.table(file = "./data2/UCI HAR Dataset/activity_labels.txt", stringsAsFactors = FALSE, col.names = c("id", "name"))

#3. fixing corresponding descriptive activity name to the values in the merged dataset as a variabe called activity name
dt_merged_filtered <- mutate(dt_merged_filtered, activity_name = as.character(factor(dt_merged_filtered$activity, labels = activity_labels$name)))
```

## Cleaning feature names and using descriptive names

The variable names are cleaned up to provide a more meaningnful representation:

```{r}
feature_select_clean <- gsub("\\(\\)", "", feature_select)
feature_select_clean <- gsub("^t(.*)$", "\\1-timedomain", feature_select_clean)
feature_select_clean <- gsub("^f(.*)$", "\\1-freqdomain", feature_select_clean)
feature_select_clean <- gsub("(Jerk|Gyro)", "-\\1", feature_select_clean)
feature_select_clean <- gsub("Mag", "-Magnitude", feature_select_clean)
feature_select_clean <- gsub("Acc", "-Acceleration", feature_select_clean)
feature_select_clean <- gsub("BodyBody", "Body", feature_select_clean)
feature_select_clean <- tolower(feature_select_clean)

#fixing variable names for the filtered dataset
names(dt_merged_filtered) <- c(feature_select_clean, "activity_name")

# displaying some of the features
print(feature_select_clean[1:20])

# displaying a sample data entry
print(head(dt_merged_filtered, 1))

```


## Creating a second, independent tidy data set with the average of each variable for each activity and each subject.

Here the observations for each subject and activity combination is grouped together and the mean is computed such that only one value is given for person 1 doing activity walking for example ( as opposed to having several entries for person 1 walking):

```{r}
#dt_tidy <- dt_merged_filtered %>% group_by(subject, activity) %>% summarise_each(funs(mean)) #%>% gather(key = measure, value = mean, -subject, -activity)

dt_tidy <- dt_merged_filtered %>% group_by(subject, activity, activity_name) %>% summarise_each(funs = list(mean))

# displaying a sample data entry
print(head(dt_tidy, 1))

```

## Saving the data and clearing up the workspace

The now tidied data is saved in the relevant workspave as a text file:
```{r}
#saving independent tidied data to a file
write.table(dt_tidy, file = "./data2/tidied_data.txt")


#Clearing the workspace
rm(list = ls())

```




